---
title: "GroupProject"
author: "Grace Rovinsky, Max Lick, Jeremy Ijoma"
date: "December 17, 2025"
format: 
  pdf:
    toc: false
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry: 
      - top=0.5in
      - left=0.5in
      - right=0.5in
      - bottom=0.5in
    colorlinks: true
execute: 
  echo: false
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

# Introduction

This document outlines our exploration into many aspects of the countries around the world. To begin, we selected the Country Data dataset found in the dcData library. This data originates from the book, Data Computing (2nd edition) by Kaplan & Beckman. To complement this dataset, we used the CountryCentroids dataset from the same source. A dataset outlining which continent each country belongs to was found and also used to assist in creating more thorough analysis. This data was found at worldpopulationreview.com and is referenced in the end of the document. Understanding how demographic, economic, and health variables interact with each other is essential for evaluating the development of countries and informing future policy. Through our exploration of these datasets, we hope to discover intriguing patterns or trends that we will then share and explain throughout the rest of the document.
We used data visualizations to identify patterns and relationships across multiple variables within our data.
By displaying multivariate data visualizations and descriptive statistics using the tidyverse and ggplot2 packages, we highlighted how complex attributes such as population growth, economic stability, and health are shaped by many connected factors.

# Summary Table

This table includes a summary of the descriptive statistics we chose to look at for this project.

```{r sum_tbl}
# Load all necessary packages
library(tidyverse) 
library(ggplot2)
library(ggpubr)
library(jpeg)
library(dcData)
library(kableExtra)
library(rvest)
library(readxl)
library(tibble)
library(dcData)
# Select key variables for summary
summary_stats <- CountryData %>%
  select(area, pop, growth, birth, death, health, obesity, educ, unemployment, life, infant, underweight) 

# Create summary table for data
summary_tbl <- psych::describe(summary_stats)
summary_tbl
```

# Population vs. Birth Rate

```{r birth_rate}
## Author: Max Lick
# Load all necessary packages
library(tidyverse) 
library(ggplot2)
library(ggpubr)
library(jpeg)
library(dcData)
library(kableExtra)
library(rvest)
library(readxl)
library(tibble)
library(dcData)
# Upload world map image
img <- readJPEG("/Users/pacad/Downloads/WorldMap.jpg")

# Select large countries
countries <- CountryData %>%
  select(c(country, birth, pop)) %>%
  filter(pop > 100000)

locations <- CountryCentroids %>%
  select(c(name, long, lat)) %>%
  rename(country = name)

dataPolished <- left_join( 
  x = countries,
  y = locations,
  by = join_by(country == country)
)

# Create Population vs. Birth Rate visualization
birth_rate <- ggplot(
  data = dataPolished,
  mapping = aes(
    x = long,
    y = lat,
    size = pop/1000000,
    color = birth
  )) + 
  background_image(img) +
  geom_point() + 
  coord_cartesian(xlim = c(-180,180), ylim = c(-90,90)) + 
  scale_color_gradient(low = 'red', high = 'blue') + 
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Map of Population and Birth Rate",
    color = "Births per 1000 People",
  ) + 
  guides(size = 'none')
birth_rate
```
This map provides a look into how population and birth rate are related from country to country. As the map shows, population does not seem to affect birth rate much, or vice versa. However, we do notice that, although some countries have larger populations, their birth rate is very similar to that of the lesser populated countries. The only areas of higher population seem to be in Africa and a small portion of the Middle East.

# Country Data by Continent

```{r ctry_by_cont}
## Author: Max Lick
# Load all necessary packages
library(tidyverse) 
library(ggplot2)
library(ggpubr)
library(jpeg)
library(dcData)
library(kableExtra)
library(rvest)
library(readxl)
library(tibble)
library(dcData)
# Read Country By Continent file
continents <- read_excel("C:/Users/pacad/Downloads/CountryByContinent.xlsx") %>%
  add_row(Brazil = "Brazil", `South America` = "South America", .before = 1) %>%
  rename(
    "country" = "Brazil",
    "continent" = "South America"
  )

countries <- CountryData

polished <- countries %>%
  select(c(1:6),12,16,18,26) %>%
  left_join(continents, by = "country") %>%
  drop_na() %>%
  rename("Continent" = "continent")

table1 <- polished %>%
  group_by(Continent) %>%
  summarise(
    Area_mean = mean(area),
    Pop_mean = mean(pop),
    Growth_mean = mean(growth),
    Birth_mean = mean(birth),
    Death_mean = mean(death)
    )

table2 <- polished %>%
  group_by(Continent) %>%
  summarise(
    Health_mean = mean(health),
    Obesity_mean = mean(obesity),
    Educ_mean = mean(educ),
    Unemployment_mean = mean(unemployment)
  )

table1 %>%
  kable( # Make Pretty Table
    caption = "Country Data Summary Statistics, Grouped By Continent", # Add Title
    align = c("l", rep("c", 6)) # Control the text alignment in each column
  ) %>%
  kable_classic( # Pre-built styling
    font_size = 12, # Control Font Size
    lightable_options = "striped" # Adds striping in a pre-built table style
  )

table2 %>%
  kable( # Make Pretty Table
    caption = "Country Data Summary Statistics, Grouped By Continent, cont'd", # Add Title
    align = c("l", rep("c", 6)) # Control the text alignment in each column
  ) %>%
  kable_classic( # Pre-built styling
    font_size = 12, # Control Font Size
    lightable_options = "striped" # Adds striping in a pre-built table style
  )
```
The above tables use the same grouping system, but had to be separated in order to fit onto the page. The tables show mean values of all countries in each continent for nine different attributes. These include area, population, growth rate, birth rate, death rate, health, obesity, education, and unemployment. These values give us an overview of the patterns occurring across the continents and can help us make assumptions at the continent level.
```
The table shows mean values of all countries in each continent for nine different attributes. These include area, population, growth rate, birth rate, death rate, health, obesity, education, and unemployment. These values give us an overview of the patterns occurring across the continents and can help us make assumptions at the continent level. Something to observe is how continents that tend to have higher population or growth rate also seem to have worse health standards.

# Health Spending, Life Expectancy, Infant Mortality

This plot illustrates the strong inverse relationship between the infant mortality rate and life expectancy of a country, incorporating health spending.
Countries that invest more heavily in healthcare tend to have lower infant mortality and longer lifespans.
This reflects real world access to medical services, emphasizing how public policy choices have a direct influence on population health outcomes.

```{r health}
# Load all necessary packages
library(tidyverse) 
library(ggplot2)
library(ggpubr)
library(jpeg)
library(dcData)
library(kableExtra)
library(rvest)
library(readxl)
library(tibble)
library(dcData)
# Create Health Spending, Infant Mortality, and Life Expectancy visualization
health_visualization <- CountryData %>%
  drop_na(infant, life, health) %>%   # Drop missing values for relevant variables
  ggplot(aes(x = infant, y = life)) +
  geom_point(aes(size = health), alpha = 0.7) +
  labs(
    title = "Health Spending and Population Health Outcomes",
    x = "Infant Mortality Rate",
    y = "Life Expectancy (years)",
    size = "Health Spending")
health_visualization
```

# GDP, Unemployment, Inflation

This plot explores macroeconomic stability by examining unemployment rates and inflation, with GDP representing the overall economic scale of a country.
Larger economies may better absorb any economic shocks, while smaller economies often face greater volatility.
High unemployment combined with high inflation can significantly reduce quality of life for a population and limit policy options.

```{r gdp}
# Load all necessary packages
library(tidyverse) 
library(ggplot2)
library(ggpubr)
library(jpeg)
library(dcData)
library(kableExtra)
library(rvest)
library(readxl)
library(tibble)
library(dcData)
# Create GDP, Inflation, and Unemployment visualization
  gdp_visualization <- CountryData %>%
  drop_na(GDP, unemployment, inflation) %>%   # Drop missing values for relevant variables
  ggplot(aes(x = unemployment, y = inflation)) +
  geom_point(aes(size = GDP), alpha = 0.7) +
  labs(
    title = "Macroeconomic Conditions by Economic Size of Country",
    x = "Unemployment Rate",
    y = "Inflation Rate",
    size = "Gross Domestic Product"
  )
  gdp_visualization
```

# FAIR and CARE Principles

## FAIR

findable: The data is easy to find.
It is taken from the 2014 CIA Factbook which can be found online.
The data is also easily accessible through the dcData package on RStudio.

accessible: The datasets are open, free, and universally implementable.

interoperable: The data uses a formal, accessible, shared, and broadly applicable language for knowledge representation.
It provides a common understanding of all attributes and countries represented in the data.

reusable: The data is thoroughly described with multiple relevant and accurate attributes.
CountryData contains 76 attributes for each country, and CountryCentroids contains 4 attributes.

## CARE

One challenge we had with determining whether or not this data conformed to the CARE principles was assessing whether or not the datasets ignored power differentials and historical contexts related to Indigenous Peoples.

collective benefit: This data includes all countries and supports use and reuse by Indigenous nations.
It enriches potential governance and engagement of citizens as it provides all nations and communities with a clear view of each country's attributes.
The data benefits all communities in an equitable manner.

authority to control: The data is available and accessible to Indigenous Peoples.
This way, the datasets support Indigenous governance.

responsibility: When using this data, we ensured that any representation of data from Indigenous nations were respectful of the dignity of all Indigenous nations and communities.

ethics: The datasets provide an accurate and fair view of all countries represented.
This minimizes harm and maximizes benefit as it does not stigmatize Indigenous communities and represents all communities fairly.

# Conclusion

Through multivariate visualizations and descriptive statistics, this exploratory data analysis reveals clear and consistent relationships among attributes of countries.
The results suggest that investments in healthcare and education are strongly associated with improved life expectancy, reduced infant mortality, and greater countrywide stability.
Macroeconomic patterns such as unemployment and inflation vary with the economic scale of a country.
Our findings do not suggest complete causation, but they generate hypotheses for potential confirmatory data analysis.
The data analysis emphasizes the value of looking into multiple variables when interpreting nationwide outcomes and establishes the role of data visualization as a powerful tool for understanding complex systems.

{{< pagebreak >}}
## Citations
Kaplan, D. (2021, January 19). Data Computing (2nd edition). Github.io. https://dtkaplan.github.io/
DataComputingEbook/

World Population Review. (2025, December 12). List of Countries by Continent. World Population Review.            https://worldpopulationreview.com/country-rankings/list-of-countries-by-continent

{{< pagebreak >}}

## Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
